---
title: "Equity Strive Capital (ESC) Risk Management"
author: "Tafara Moyo"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

## 1. Introduction

The project aims to develop a Value-at-Risk model that estimates the
maximum potential loss in a portfolio with equal investment in top 5
performing stocks from 2020 to date.The major is is to try to see if the
current upward movement in stock prices is sustainable or just the
beginning of th end and if the portfolio is at risk of losing value.
This study will assume a a factious portfolio called Equity Strive
Capital. The portfolio will be made up of the following stocks: Google
(GOOGL), Apple (AAPL), Microsoft (MSFT), Amazon (AMZN), and Nvidia
(NVDA).

## 2. Rational for the study

The project is inspired by the following needs:

**Market Analysis:** Investigate whether the recent upward trend in
stock prices, particularly in the technology sector, represents a stable
growth opportunity or signals an impending downturn. Can an all tech
portfolio be profitable from this?

***Risk Management:*** Utilize VaR to quantify potential losses in the
portfolio, aiding in informed decision-making about risk exposure amidst
market volatility.

## 3. Objectives

The objectives of this study are as follows:

1.  To develop a VaR model from 5 best performing tech stocks

2.  To come up with insights on how to manage risk in 2024.

## 4. Methods

The study will be conducted in the following steps:

1.  Data Collection: Gather historical price data for 5 top-performing
    assets.

2.  Data Pre-processing: Clean and preprocess both financial and
    sentiment data, ensuring alignment for analysis.

3.  VaR Model Development: Explore and implement various VaR methods
    historical simulation, Monte Carlo simulations.

4.  Insights and Recommendations: Provide insights and recommendations
    based on the VaR model results.

## 5. Set Up

------------------------------------------------------------------------

The following libraries will be used for the project:

```{r}
suppressMessages({
  library(quantmod)
  library(TTR)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(glue)
  library(httr)
  library(RJSONIO)
  library(purrr)
  library(reshape2)
})
```

### Data Collection

------------------------------------------------------------------------

**Collecting Stock prices data**

The data will be collected from Alpha Vantage API, which provides
historical stock prices and other financial data. The following code
fetches weekly stock prices for the top-performing assets (e.g., GOOGL,
AAPL, MSFT, AMZN, NVDA) using the AlphaVantage API.These stocs are part
of the magnificent seven stocks and offer very good returns in the
market and are very liquid.

```{r}
##Fetching the top performing stocks
fetch_top_performing_stocks <- function(api_key, symbol) {
 base_url <- "https://www.alphavantage.co/query?function=TIME_SERIES_WEEKLY&symbol={symbol}&apikey={api_key}"
  url <- glue(base_url, api_key = api_key, symbol = symbol)
  response <- httr::GET(url[1])
  content <- httr::content(response, "text")
  data <- RJSONIO::fromJSON(content)
 
  if (is.null(data$`Weekly Time Series`)) {
    return(data.frame(Date = as.Date(character()), Weekly_Return = numeric(), Symbol = symbol))
  }
 
  weekly_series <- data$`Weekly Time Series`
  dates <- names(weekly_series)
  prices <- purrr::map_dbl(dates, ~as.numeric(weekly_series[[.x]][4]))
  returns <- c(NA, diff(log(prices)))
 
  data.frame(Date = as.Date(dates), Weekly_Return = returns, Symbol = symbol)
 
}

api_key <- "1U1APZAOHQ6NU9YS"
symbols <- c("GOOGL", "AAPL", "MSFT", "AMZN", "NVDA")

output <- lapply(symbols, function(x) fetch_top_performing_stocks(api_key, x))
```

Since the API call is limited, l will save the data into csv files for
future use. Avoiding to exhaust the 25 daily cap.

```{r}
##Saving my data into seperate dataframes
for (i in 1:length(output)) {
  df <- output[[i]]
  symbol <- df$Symbol[1]
  filename <- paste(symbol, "weekly_returns.csv", sep = "_")
  write.csv(df, filename, row.names = FALSE)
}

```

***Reloading the data from the csv files***

```{r}
GOOGL <- read.csv("~/Downloads/Stocks/GOOGL_weekly_returns.csv")
AAPL <- read.csv("~/Downloads/Stocks/AAPL_weekly_returns.csv")
MSFT <- read.csv("~/Downloads/Stocks/MSFT_weekly_returns.csv")
AMZN <- read.csv("~/Downloads/Stocks/AMZN_weekly_returns.csv")
NVDA <- read.csv("~/Downloads/Stocks/NVDA_weekly_returns.csv")
```

### Data Pre-processing

Here l want to check the structure of my data and see if l need to make
any changes to it. I will also combine my data into a portfolio return
data frame and then calculate the weekly returns to enable further
calculations.

```{r}
#Just checking the structure of my data to see if l need to make any changes
str(GOOGL)
str(AAPL)
str(MSFT)
str(AMZN)
str(NVDA)
```

```{r}

# Combining data frames and ensuring 'Date' column is in Date format
all_stocks <- bind_rows(GOOGL, AAPL, MSFT, AMZN, NVDA) %>%
  mutate(Date = as.Date(Date)) %>%
  filter(Date >= as.Date("2020-01-01")) 

# Removing rows with NA values (assuming you want to check all columns)
all_stocks <- na.omit(all_stocks)

#### Combining this into one Data Frame of portfolio return

portfolio_weekly_returns <- all_stocks %>%
  group_by(Date) %>%
  summarise(Weekly_Return = mean(Weekly_Return, na.rm = TRUE))
#Now l have a data frame for the eman return of the portfolio from 2020 
```

## 6. Calculations

**Value at Risk (VaR)**

Value at Risk (VaR) is a risk management tool that calculates, at a
particular confidence level, the maximum expected loss of an investment
portfolio over a specific period of time.The query "What is the maximum
amount I could lose with a given probability over a certain period?" is
answered by it. VaR is well-liked because it makes risk exposure to both
technical and non-technical stakeholders clear and simple to understand.
The following code calculates VaR for Stewards Portfolio

```{r}
confidence_level <- 0.05  # 95% confidence level

# Calculate Historical VaR
historical_VaR_portfolio <- quantile(portfolio_weekly_returns$Weekly_Return, confidence_level, na.rm = TRUE)

historical_VaR_portfolio
```

There is a 95% confidence level that the maximum potential loss over the
specified time period will not exceed -7.36% of the portfolio. In
simpler terms, it suggests that in 95 out of 100 scenarios, the
portfolio's loss would not be worse than 7.36%, reflecting the
portfolio's risk exposure under normal market conditions.

**Monte Carlo Simulation**

Monte Carlo Simulation is used to simulate the likelihood of various
outcomes in a process that is difficult to predict because of the
influence of random factors. It is extensively utilized for risk
assessment using random sampling to produce numerical data, the method
entails running a large number of simulations to generate alternative
outcomes for a random process.

```{r}
set.seed(123)  # For reproducibility
n_simulations <- 10000
n_days <- 5  # Assuming 5 trading days in a week

mean_return <- mean(portfolio_weekly_returns$Weekly_Return, na.rm = TRUE)
std_return <- sd(portfolio_weekly_returns$Weekly_Return, na.rm = TRUE)

# Simulate weekly returns
simulated_returns <- replicate(n_simulations, sum(rnorm(n_days, mean_return / n_days, std_return / sqrt(n_days))))

# Calculate Monte Carlo VaR
monte_carlo_VaR_portfolio <- quantile(simulated_returns, confidence_level)
print(monte_carlo_VaR_portfolio)
```

Therefore, the Value at Risk (VaR) calculation indicates that, with a
95% confidence level, the maximum expected loss for the Stewards
Portfolio over the specified period is -12.29%. This means there is a
95% probability that the portfolio will not lose more than 12.29% of its
value in the given timeframe, according to the scenarios generated by
the Monte Carlo simulation.

**VaR Comparison Visualization**

```{r}
# Combine VaR values into a data frame for plotting
var_values <- data.frame(Type = c("Historical VaR", "Monte Carlo VaR"),
                         VaR = c(historical_VaR_portfolio, monte_carlo_VaR_portfolio))

ggplot(var_values, aes(x = Type, y = VaR, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Portfolio Value at Risk (VaR) Since 2020", x = "", y = "Value at Risk") +
  scale_fill_brewer(palette = "Pastel1")

```

From the above chart, we can see that the Monte Carlo VaR estimate is
higher than the Historical VaR estimate, indicating a higher potential
loss for the portfolio. This suggests that the Monte Carlo simulation,
which accounts for a wider range of potential outcomes, is more
conservative in its risk assessment compared to the Historical VaR
method.

------------------------------------------------------------------------

**Portfolio Performance Visualization**

The following visualizations will be used to assess the performance of
the Equity Strive Capital Portfolio since 2020.

```{r}

ggplot(portfolio_weekly_returns, aes(x = Date, y = Weekly_Return)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Portfolio Weekly Returns Over Time", x = "Date", y = "Weekly Return (%)") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```

We can see that we the portfolio has been having more losses than gains
since 2020. This is a clear indication that the portfolio is at risk of
losing value.

**Cumulative Returns Plot**

```{r}

portfolio_weekly_returns$Adjusted_Return = portfolio_weekly_returns$Weekly_Return / 100 + 1

# Calculate Cumulative Return
portfolio_weekly_returns$Cumulative_Return = cumprod(portfolio_weekly_returns$Adjusted_Return)

ggplot(portfolio_weekly_returns, aes(x = Date, y = Cumulative_Return)) +
  geom_line(color = "red") + # Change to "blue" for a blue line
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Portfolio Cumulative Returns Over Time", x = "Date", y = "Cumulative Return (%)") +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "black", colour = "black"),
        plot.background = element_rect(fill = "black", colour = "black"),
        text = element_text(color = "white"),
        axis.text = element_text(color = "white"),
        axis.title = element_text(color = "white"),
        panel.grid.major = element_blank(),  # Remove major gridlines
        panel.grid.minor = element_blank(),  # Remove minor gridlines
        plot.title = element_text(hjust = 0.5, color = "white"))


```

The portolio has been having cumulative gains in 2023 but this has been
drastically falling since about mid-2023. This is a clear indication
that the portfolio is at risk of losing value.

**Correlation Analysis**

```{r}

returns_by_symbol <- all_stocks %>%
  select(Date, Symbol, Weekly_Return) %>%
  spread(Symbol, Weekly_Return)

cor_matrix <- cor(returns_by_symbol[, -1], use = "complete.obs")

ggplot(data = melt(cor_matrix), aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Correlation Matrix of Portfolio Stocks", x = "Stock", y = "Stock") +
  theme_minimal()

```

Most of these stocks belong in the consumer technology sector and are
highly correlated. This means that they are likely to move in the same
direction. This is a clear indication that the portfolio is at risk of
losing value and confirms the fall in returns shown before.

## 7. Conclusions

**1. Portfolio Diversification:** To mitigate risk associated with the
uniform direction of tech stock movements, diversify the portfolio by
incorporating non-tech sectors\
**2. Alternative data:** Leverage alternative data sources to uncover
unique investment opportunities and enhance portfolio decision-making
beyond traditional market indicators.

## 8. Resources

For further understanding of these concepts, the following resources
were used:

1\. [Why Log Returns -
Quantivity](https://quantivity.wordpress.com/2011/02/21/why-log-returns/)

2\. [Value at Risk -
Investopedia](http://www.investopedia.com/terms/v/var.asp)

3\. [Approaches to VaR -
Stanford](https://web.stanford.edu/class/msande444/2012/MS&E444_2012_Group2a.pdf)

4\. [Uniform Distribution -
R](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Uniform.html)

5\. [Monte Carlo Method in R - alookanalytics
blog](https://blog.alookanalytics.com/2017/04/26/monte-carlo-method-in-r/)

6\. [Calculating VaR with R -
R-Bloggers](https://www.r-bloggers.com/calculating-var-with-r/)

7\. [Monte Carlo Package -
R](https://cran.r-project.org/web/packages/MonteCarlo/MonteCarlo.pdf)

8\. [Fixed Income Risk: Calculating Value at Risk (VaR) for
Bonds](https://financetrainingcourse.com/education/2013/05/bond-risk-calculating-value-at-risk-var-for-bonds/)

9\. [Portfolio & Risk Analytics - Bloomberg
Terminal](https://www.bloomberg.com/professional/product/portfolio-risk-analytics/)

10\. [Risk Management for Fixed Income Portfolios - Credit
Suisse](https://www.credit-suisse.com/media/am/docs/asset_management/events/2014/fits2014-program/3-5-2-traband-risk-management.pdf)

11\. [Fixed Income VaR - Krzysztof
Ostaszewski](https://math.illinoisstate.edu/krzysio/MAT483/ValueAtRisk.pdf)

12\. [Manipulating Time Series Data in R with xts & zoo -
RPubs](https://rpubs.com/mohammadshadan/288218)
```{r}

```

